# GAM Visualization Scripts - Quick Start Guide

## Installation

```{r}
# Required packages
# install.packages(c("ggplot2", "dplyr", "tidyr", "viridis", 
#                    "patchwork", "scales", "mgcv", "gratia"))

library(here)
# Load the scripts
source(here("R scripts", "gam_NEE_vizfunctions.R"))
```

## Basic Usage

### 1. Load Your Data

```{r}
library(readr)
library(lubridate)

# Load your EC daily management data
data <- read_csv(here("eddy_covariance_fluxdata", "ecdailymngmnt_16Mar2026.csv"))
data$date <- as.Date(data$date, format="%m/%d/%y")
```

### 2. Fit a GAM Model
```{r}
library(mgcv)

# make sure management is in factor form
data$management <- factor(as.character(data$management))
# same for crop stage:
data$crop_stage_simple <- factor(
  data$crop_stage_simple,
  levels = c("mature", "early", "vegetative", "reproductive", "grain_fill", "dormant", "fallow", "regrowth", "recovery")
)

# Example: NEE model with crop stage interaction
model_nee <- gam(
  NEE ~ 
    crop_stage_simple +
    s(air_temperature, by = crop_stage_simple, k = 5) +
    s(VPD, k = 5) +
    # s(soil_temp_smooth, k = 5) +
    management,
  data = data,
  method = "REML"
)

summary(model_nee)
```

### 3. Create Visualizations

#### Quick Single Plot
```{r}
# Observed vs predicted
plot_gam_observed_vs_predicted(model_nee, data, "NEE")
```


```{r}
# Residuals by stage
plot_gam_residuals_by_stage(model_nee, data, "NEE")
```


```{r}
# Flux distribution by stage
plot_flux_by_stage(data, "NEE", add_stats = TRUE)
```

#### Save All Standard Plots
```{r}
# Generate and save all plots automatically
# generate_all_plots(data, model_nee, output_dir = "./results/plots")

```

## Individual Plot Examples

### A. Model Performance
```{r}
# Observed vs predicted with hexbin
p <- plot_gam_observed_vs_predicted(
  model = model_nee,
  data = data,
  flux_var = "NEE",
  title = "GAM Performance: Net Ecosystem Exchange"
)
print(p)
# ggsave("obs_vs_pred.png", p, width = 8, height = 6, dpi = 300)
```

### B. Stage Comparisons
```{r}
# Compare flux across all stages
p <- plot_flux_by_stage(
  data = data,
  flux_var = "NEE",
  add_stats = TRUE,
  facet_by = NULL  # Or "management" to split by field
)
print(p)
```

### C. Management Comparison
```{r}
# Conventional vs Organic
p <- plot_management_comparison(
  data = data,
  flux_var = "NEE"
)
print(p)
```

### D. Time Series
```{r}
# Full time series with stage coloring
p <- plot_timeseries_with_stages(
  data = data,
  flux_var = "NEE",
  management = "organic",  # Optional: filter to one field
  year = 2021              # Optional: filter to one year
)
print(p)
```

### E. Partial Effects
```{r}
# Temperature effect across stages
p <- plot_partial_effects_by_stage(
  model = model_nee,
  data = data,
  var = "air_temperature",
  flux_var = "nee",
  stages = c("vegetative", "reproductive", "grain_fill", "mature")
)
print(p)
```

### F. Heatmap
```{r}
# Seasonal patterns
p <- plot_flux_heatmap(
  data = data,
  flux_var = "NEE",
  time_unit = "month",
  management = "conventional"
)
print(p)
```

### G. Multi-Panel Summary
```{r}
# All three flux components (NEE, GPP, Reco)
p <- plot_flux_summary(
  data = data,
  stages = NULL,  # All stages
  management = NULL  # Both fields
)
print(p)
# ggsave("flux_summary.png", p, width = 10, height = 12, dpi = 300)
```

## Advanced Examples

### Compare Multiple Flux Variables
```{r}
library(patchwork)

# Create plots for each flux
p_nee <- plot_flux_by_stage(data, "NEE")
p_gpp <- plot_flux_by_stage(data, "GPP")
p_reco <- plot_flux_by_stage(data, "Reco")

# Combine
combined <- p_nee + p_gpp + p_reco +
  plot_layout(ncol = 3) +
  plot_annotation(title = "Carbon Flux Components by Crop Stage")
combined

# ggsave("flux_comparison.png", combined, width = 18, height = 6, dpi = 300)
```

### Facet by Year and Management
```{r}
# Add year column
data$year <- year(data$date)

# Plot with faceting
p <- ggplot(data, aes(x = crop_stage_simple, y = NEE, 
                      fill = crop_stage_simple)) +
  geom_boxplot(alpha = 0.8) +
  facet_grid(management ~ year) +
  scale_fill_manual(values = stage_colors, guide = "none") +
  theme_flux() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "NEE by Stage, Management, and Year",
       x = "Crop Stage",
       y = expression(NEE~(µmol~CO[2]~m^-2~s^-1)))

print(p)
```

### Smooth Term Visualization
```{r}
library(gratia)

# Plot smooth effects
draw(model_nee, select = "s(air_temperature):crop_stage_simple", partial_match = TRUE)
draw(model_nee, select = "s(VPD)", partial_match = TRUE)
```

### Recovery vs Regrowth Comparison (Alfalfa)
```{r}
# Filter to alfalfa and post-harvest stages
alfalfa_data <- data %>%
  filter(current_crop == "alfalfa hay",
         crop_stage_simple %in% c("recovery", "regrowth", "vegetative"))

# Compare fluxes
p <- ggplot(alfalfa_data, aes(x = crop_stage_simple, y = NEE_gapfilled,
                              fill = crop_stage_simple)) +
  geom_violin(alpha = 0.6, draw_quantiles = 0.5) +
  geom_jitter(alpha = 0.3, width = 0.2) +
  scale_fill_manual(values = stage_colors[c("recovery", "regrowth", "vegetative")],
                    guide = "none") +
  facet_wrap(~ last_harvest_same_crop, labeller = "label_both") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_flux() +
  labs(title = "Alfalfa Post-Harvest Progression",
       subtitle = "Recovery (days 0-5) vs Regrowth (days 5-15) vs Vegetative",
       x = "Stage",
       y = expression(NEE~(µmol~CO[2]~m^-2~s^-1)))

print(p)
```

### Custom Color Schemes
```{r}

# Modify the default colors
stage_colors_custom <- c(
  "fallow" = "#E8E8E8",
  "recovery" = "#D32F2F",
  "early" = "#81C784",
  "regrowth" = "#4CAF50",
  "vegetative" = "#2196F3",
  "reproductive" = "#FF4081",
  "grain_fill" = "#FF9800",
  "mature" = "#795548",
  "dormant" = "#9E9E9E"
)

# Use in plots
p <- plot_flux_by_stage(data, "NEE") +
  scale_fill_manual(values = stage_colors_custom, guide = "none")
p
```

## Tips and Best Practices

### 1. Check Data Distribution
```{r}
# Before modeling, check sample sizes
table(data$crop_stage_simple)
table(data$crop_stage_simple, data$management)
```

### 2. Handle Outliers
```{r}
# Option 1: Filter extreme values
data_filtered <- data %>%
  filter(NEE > quantile(NEE, 0.01, na.rm = TRUE),
         NEE < quantile(NEE, 0.99, na.rm = TRUE))

# Option 2: Winsorize
library(DescTools)
data$NEE_winsorized <- Winsorize(data$NEE, val = c(0.01, 0.99))
```

### 3. Model Diagnostics
```{r}
# Use gratia for comprehensive diagnostics
library(gratia)

appraise(model_nee)

# Check concurvity
concrvity(model_nee)

# Check basis dimensions
gam.check(model_nee)
```

### 4. Compare Models
```{r}
# Fit multiple models
model1 <- gam(NEE ~ crop_stage_simple + s(air_temperature), data = data)
model2 <- gam(NEE ~ crop_stage_simple + s(air_temperature, by = crop_stage_simple), 
              data = data)

# Compare AIC
AIC(model1, model2)

# ANOVA comparison
anova(model1, model2, test = "F")
```

### 5. Publication-Ready Exports
```{r}
# High-resolution for publication
ggsave("figure1.png", plot, width = 7, height = 5, dpi = 600)
ggsave("figure1.pdf", plot, width = 7, height = 5)  # Vector format
ggsave("figure1.tiff", plot, width = 7, height = 5, dpi = 300, compression = "lzw")
```

## Troubleshooting

### Issue: "Object not found" errors
```{r}
# Make sure all required columns exist
required_cols <- c("date", "NEE_gapfilled", "GPP_gapfilled", "Reco_gapfilled",
                   "crop_stage_simple", "management", "air_temperature")
missing <- setdiff(required_cols, names(data))
if (length(missing) > 0) {
  stop("Missing columns: ", paste(missing, collapse = ", "))
}
```

### Issue: Plots look crowded
```{r}

# Filter to specific stages or time periods
data_subset <- data %>%
  filter(crop_stage_simple %in% c("vegetative", "reproductive", "mature"),
         year(date) == 2021)

plot_flux_by_stage(data_subset, "NEE_gapfilled")
```

### Issue: GAM won't converge
```{r}
# Reduce basis dimension
model <- gam(NEE ~ s(air_temperature, k = 4), data = data)  # Lower k

# Or use different optimizer
model <- gam(NEE ~ s(air_temperature), data = data, optimizer = "optim")
```

## Additional Resources
```{r}
# -   **mgcv documentation:** `?gam`, `?s`, `?ti`
# -   **gratia package:** For modern GAM visualization
# -   **GAM tutorial:** Wood (2017) - Generalized Additive Models: An Introduction with R
# -   **Color palettes:** viridis, RColorBrewer, ggsci packages

## Contact

For questions or issues with these scripts, check: 1. Your data structure matches expected format 2. All required packages are installed 3. GAM model converged successfully 4. Column names match those used in functions
