---
title: "eddycovariance_HC"
author: "Elizabeth Forbes"
format: html
editor: visual
execute:
  echo: true       # shows code
  warning: false    # hides warnings
  message: false    # hides messages
  error: false      # hides errors (optional, use carefully)
  output: true     # standard output (like print results)
---

# Gas exchange across agriculture management strategy

This study examines the differences in average and cumulative greenhouse gas exchange from two different fields located in the Hudson Valley (Columbia County). Both fields are grain fields growing primarily corn and soy, with occasional cover cropping and rotations of barley. The main difference in farming strategy is organic certification, with one field being organic and the other conventionally-operated. However they share many practices, including tillage, fertilizing, cover cropping. However they each have stochastic management events not shared by the other, including herbicide use (conventional) and growing alfalfa hay (organic).

Data were collected from 2018-2022 at two eddy covariance towers equipped with carbon dioxide sensors, one located in each field. Between 2019 and 2020, data were also collected using handheld survey gas exchange units (for both soil respiration and NEP). (Survey data were collected in more years, but these data are not yet cleaned and analyzable.) Other data sources, such as photosynthetically active radiation (PAR); management records; and soil organic matter (SOM) were also collected for subsets of the period during which the towers were deployed and collecting data.

The purposes of these analyses include demonstrating the difference in average and cumulative gas exchange across treatments (organic vs. conventional) using eddy covariance data, augmented and supported by survey data and meta-data where necessary and/or appropriate.

## Libraries

```{r libraries}
library(tidyverse)
library(readxl)
library(readr)
library(ggpubr)
library(naniar)
library(here)
library(sf) # for reading shape files
library(purrr)
library(lubridate)
library(hms)
library(ineq) # for Lc function: lorenz curve
library(ggstance)
library(visibly)
library(mgcv) # visualizing gam model results in Dharma
library(DHARMa)
library(gratia)
library(zoo)
library(epiR)  # For Lin's CCC calculation
library(GGally)
library(corrplot)
library(emmeans)
library(here)
```

## Upload cleaned data to global environment

### Eddy covariance, survey, PAR/weather at each tower, and management data:

```{r data_upload}
#| warning: false

source(here("R scripts", "datacleaning_EC.r"))
source(here("R scripts", "datacleaning_chambers.r"))
# source(here("R scripts", "datacleaning_PAR_attowers.r"))
source(here("R scripts", "datacleaning_management.r"))

```

### Add management data to the EC, chamber data:

Using a custom function, calculate continuous variables for each management action. The custom function is informed by characteristics like the expected crop-specific phenology, and days since actions like tillage.

```{r managementdatamerge}
source(here("R scripts", "merge_management_data_covercropsandperennials_v3_6_0_6.r"))

ec_daily_mngmnt <- create_management_vars_multicrop(
  ec_daily_avg,
  mngmnt) 

# pull the assigned crop stages from the above to the daily chamber data so that the crop stages match across the two datasets:
chamber_daily_mngmnt <- chamber_daily_site %>%
  # convert date from POSIXct to Date:
  mutate(date = as.Date(date, tz = "EST")) %>%
  left_join(ec_daily_mngmnt %>% 
              select(management, date, current_crop, crop_stage, crop_stage_simple),
            by = c("management", "date"))

```

Plot conventional, un-gapfilled Reco with tillage events:

```{r Reco_withtillageevents_conv}

ec_daily_mngmnt %>% 
  filter(management == "conventional") %>% 
  ggplot(aes(x=date, y=Reco))+
  geom_point(alpha = 0.25, color = "red")+
  geom_line(alpha = 0.25)+
  geom_point(aes(x=date, y=tillage_event), fill = "blue",
             size = 4,shape = 21, color = "black", alpha = 0.5)+
  theme_bw()+
  ggtitle("Conventional grain field ecosystem respiration with noted tillage events")+
  ylab(expression(paste("ecosystem respiration, " * mu * "mol m"^-2 * " s"^-1 *"")))+
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

Plot organic, un-gapfilled Reco with tillage events

```{r Reco_withtillageevents_org}

ec_daily_mngmnt %>% 
  filter(management == "organic") %>% 
  ggplot(aes(x=date, y=Reco))+
  geom_point(alpha = 0.25, color = "darkgreen")+
  geom_line(alpha = 0.25)+
  geom_point(aes(x=date, y=tillage_event), fill = "blue", 
             size = 4,shape = 21, color = "black", alpha = 0.5)+
  theme_bw()+
  ggtitle("Organic grain field ecosystem respiration with noted tillage events")+
  ylab(expression(paste("ecosystem respiration, " * mu * "mol m"^-2 * " s"^-1 *"")))+
  theme(axis.title.x = element_blank(),
        legend.position = "none")

```

### Gantt chart of crop rotations:

```{r}

mngmnt$duration <- as.numeric(mngmnt$duration)

mngmnt %>% 
  ggplot(aes(x = date, y = management, color = crop)) +
  geom_point(size = 5, alpha = 0.75,
             position = ggstance::position_dodgev(height = 0.7)) +
  scale_x_date(date_labels = "%d %b %Y") +
  theme_minimal() +
  labs(title = "Crop Rotations: Organic and Conventional Grains",
       x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_color_brewer(palette = "RdYlBu")
```

### Gantt chart of other management activities:

```{r}
mngmnt %>% 
  ggplot(aes(x = date, y = management, color = category)) +
  geom_point(size = 5, alpha = 0.75,
             position = ggstance::position_dodgev(height = 0.7)) +
  scale_x_date(date_labels = "%d %b %Y") +
  theme_minimal() +
  labs(title = "Farm Management Activities Timeline",
       x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_color_brewer(palette = "RdYlBu")
```

# Analyses

## GAM: EC data, daily averages

Assessing daily average trends is useful because it gives mechanistic insight into how gas exchange varies across management types. This analysis tells us when and how management moderates fluxes throughout the seasons we measured.

### NEE

#### GAM model development: daily average

```{r corplots}
#| warning: false

ec_daily_mngmnt %>% 
  ggpairs(columns = c("VPD", 
                      "ndvi_era5", 
                      "RH", 
                      # "precip", 
                      "air_temperature", 
                      # "soilt_avg", 
                      # "vwc_avg", 
                      "days_since_tillage"),
          progress = FALSE) # removes extensive printing

# are any variables co-varying enough to warrant removal from contention for model testing? Potentially worth removing relative humidity, given that VPD is calculated with air temperature and RH. 
```

Model testing, average NEE over time: simple with main effects, then with different responses by management, then with replacement of NDVI with crop stage to account for phenology. resource on GAMs: https://deepwiki.com/cran/mgcv/2.1-the-gam()-function resource on smoothing and interaction variables: https://deepwiki.com/cran/mgcv/3.1-types-of-smooths

```{r NEE_model}
library(mgcv)

# make sure management is in factor form
ec_daily_mngmnt$management <- factor(as.character(ec_daily_mngmnt$management))
# same for crop stage:
ec_daily_mngmnt$crop_stage_simple <- factor(
  ec_daily_mngmnt$crop_stage_simple,
  levels = c("mature", "early", "vegetative", "reproductive", "grain_fill", "dormant", "fallow")
)

# examine main effects first
g1 <- gam(NEE ~ 
            management +
            s(year, bs = "re") + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern
            
            s(soilt_avg, k = 10) +
            s(VPD, k = 10) +
            s(vwc_avg, k = 10) +
            s(ndvi, k = 10) +
            # s(precip, k = 8) +
            s(days_since_tillage, k = 8),
            # s(days_since_planting, k = 8),
            
          data = ec_daily_mngmnt4,
          method = "REML")

# allow different responses by management: this is essentially an interaction of management for each term, which makes sense as we might expect each of the variables below to vary with management (soil temp and moisture, greenness, vapor pressure density)
g2 <- gam(NEE ~ 
            management +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(ndvi, by = as.factor(management), k = 10) +
            # s(precip, by = as.factor(management), k = 8) +
            s(days_since_tillage, by = as.factor(management), k = 8),
            # s(days_since_planting, by = as.factor(management), k = 8),
            
          data = ec_daily_mngmnt4,
          method = "REML")

# add vwc/soil temp interaction:
g2.2 <- gam(NEE ~ 
            management +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern
            
            ti(soilt_avg, vwc_avg, by = as.factor(management), k = c(8, 8)) +
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(ndvi, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# replace NDVI with crop phase (simple):
g3 <- gam(NEE ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# replace NDVI with crop phase (simple) plus interaction of vwc and soil temp:
g3.2 <- gam(NEE ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
            
            ti(soilt_avg, vwc_avg, by = as.factor(management), k = c(8, 8)) +
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# add interaction term between crop stage and management:
g3.3 <- gam(NEE ~ 
            management * # interaction
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
              
            ti(soilt_avg_era5, vwc_avg_era5, by = as.factor(management), k = c(8, 8)) +
            
            s(soilt_avg_era5, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg_era5, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          # data = ec_daily_mngmnt4,          
          data = ec_daily_mngmnt,

          method = "REML")

# compare using AIC values:
AIC(g1, g2, g2.2, g3, g3.2, g3.3)
# AIC(g3, g3.2, g3.3)

      #   df       AIC
# g1	  41.32714	7870.767		
# g2	  74.20071	7463.932		
# g2.2	96.54754	7430.443		
# g3	  66.30796	7387.526		
# g3.2	84.24336	7346.628		
# g3.3	89.82250	7338.887		
```

```{r NEE_gam_eval}

# improvement between 3 and 3.2 (addition of interaction term between soil temp, moisture)
anova(g3, g3.2, test = "Chisq")
#   Resid. Df Resid. Dev     Df Deviance Pr(>Chi)
# 1    1533.8     8594.0                              
# 2    1511.8     8193.1 22.038    400.9 1.079e-07 ***

# improvement between model 3 (no interaction between management and crop stage) and 3.3 (yes interaction)
anova(g3.2, g3.3, test = "Chisq")
#   Resid. Df Resid. Dev     Df Deviance  Pr(>Chi)    
# 1    1511.8     8193.1                           
# 2    1505.7     8097.3 6.058   95.788 0.006318 **
```

Going to use simple crop phase as a categorical response, rather than NDVI as a smoothing response, to investigate the effects of management on the different life stages of the crops across the time series.

```{r NEE_gam_modviz}
# Check diagnostics of all models with simple crop stage as variable, not NDVI
# appraise(g3)
# appraise(g3.2)
appraise(g3.3)
# crop stage because it is a simpler variable than NDVI, and is attuned to the different crops being planted in the rotation at a given time, as well as cover cropping.
```

significance of smooth terms:

```{r NEE_gam_output}

plot(g3.3, pages = 1)

```

```{r}
gam.check(g3.3)
summary(g3.3) 

 # for parametric terms: a significant coefficient means that the variable has a meaningful linear effect on the response.

# for smoothing terms:
#   edf (Estimated degrees of freedom): Reflects the complexity of the smooth.
# Close to 1 means near-linear.
# Higher values mean more flexible, wiggly functions.
# F-value and p-value: Tests whether the smooth term significantly improves the model.

# model fit stats:
# R-sq.(adj) =  0.563   Deviance explained = 58.5%
# -REML = 3707.1  Scale est. = 5.3019    n = 1607

```

##### Interpretation, linear variables:

Note: baseline comparison to conventional management and mature crop stage (estimated NEE = -0.2773)

There is a significant effect of organic management on average daily NEE in the mature crop stage, reducing it by 2.5 units when holding other factors constant; this makes sense, in that we've seen greater drawdown of ecosystem CO2 in the organic field compared to the conventional one when comparing them directly.

There are both trends and significant effects of individual crop stage relative to the baseline estimate (conventional, mature). Compared to the conventional mature, the early phase is higher by 1 unit; the reproductive phase is lower by 1.9 units; the grain-fill phase is significantly lower; the fallow is significantly higher.

There are also significant interactions between management and crop stage, which are the difference in the crop stage effect between organic and conventional relative to the baseline crop stage; the interaction coefficient therefore tells us how much more or less the difference between that crop stage and the baseline changes under organic compared to conventional management. (To calculate the absolute organic vs. conventional difference in NEE for each crop phase, we'd combine the main effect of organic + the interaction effect of organic on that phase.)

The effect of organic management on the influence of the vegetative phase is -1.6; the effect of organic management on the influence of the grain fill stage compared to conventional mature is -1.9. Slight interaction between organic and dormant (-1.9).

##### Interpretation, smoothing variables:

Almost every single smoothing factor is significant in driving NEE, with the exception of VWC (soil moisture) in the conventional field, demonstrating the effect of each crop stage. The F-test and p-values indicate with the smooth term explains a meaningful amount of the variation in the model's ability to predict the data. The EDF (estimated degrees of freedom) indicate the complexity of the smoothing term: close to 1 means close to linear, and larger indicates a very 'wiggly' or nonlinear relationship between the smoothing term and NEE.

Seasonality (doy, day of year) has a significant non-linear effect on NEE, capturing expected seasonal variation. The tensor interaction between soil temperature and moisture indicates a significant non-linear effect of the two as they are combined, the extent to which differs by management -- the combined effect is important for NEE, but the estimated complexity of this effect is greater in the organic field than the conventional one. This indicates that there are more nuanced and/or variable effects of this interaction there, perhaps due to more nuanced soil properties, soil microbiome, plant-soil interactions, etc. For soil temp and moisture individually, as well as VPD, the nonlinearity of effect is more pronounced in organic than conventional management, though soil moisture does not significantly moderate NEE in the conventional field at all. Days since tillage are similar.

##### Takeaways:

Organic management creates a tendency toward lower NEE, implying either less CO2 respiration or more uptake via photosynthesis, and/or a combination of the two. Crop stage modifies the effect significantly as well, indicating a different effect of crop stage on NEE depending on management strategy, but overall organic management reduces NEE. The smoothing terms universally indicated nonlinear effects on the organic field's NEE, and almost all except for soil moisture for the conventional field; for the conventional field, the effects were less nonlinear all around.

Interestingly, the random effect of year demonstrates a) that there IS a significant difference in NEE across years, but also that b) these differences are also dependent on management type, indicating that NEE reacts differently to yearly weather patterns, management change, etc., in each field.

##### Visualize:

###### Interactions, NEE daily average:

```{r GAM_NEE_interactionsplot}
# Create new data frame for prediction
newdata <- expand.grid(
  management = c("conventional", "organic"),
  crop_stage_simple = levels(ec_daily_mngmnt2$crop_stage_simple),
  year = 2019,          # hold other variables constant at mean or typical values
  doy = median(ec_daily_mngmnt2$doy),
  soilt_avg = mean(ec_daily_mngmnt2$soilt_avg),
  vwc_avg = mean(ec_daily_mngmnt2$vwc_avg),
  VPD = mean(ec_daily_mngmnt2$VPD, na.rm = TRUE),
  days_since_tillage = mean(ec_daily_mngmnt2$days_since_tillage, na.rm = TRUE)
)

# Predict fitted values with standard errors
pred <- predict(g3.3, newdata, se.fit = TRUE, type = "response")

newdata <- newdata %>%
  mutate(
    fit = pred$fit,
    se = pred$se.fit,
    lower = fit - 1.96 * se,
    upper = fit + 1.96 * se
  )

# Plot predicted NEE by crop stage and management
ggplot(newdata, aes(x = crop_stage_simple, y = fit, color = management, group = management)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_line(position = position_dodge(width = 0.5), size = 1) +
  labs(
    title = "Predicted NEE by Crop Stage and Management Type",
    x = "Crop Stage",
    y = "Predicted NEE",
    color = "Management"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Reco

#### GAM model development: daily average

```{r Reco_model}
# Again, going with crop phase over NDVI, so skipping that part of model development.

r1 <- gam(Reco ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# plus interaction of vwc and soil temp:
r1.2 <- gam(Reco ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
            
            ti(soilt_avg, vwc_avg, by = as.factor(management), k = c(8, 8)) +
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# add interaction term between crop stage and management:
r1.3 <- gam(Reco ~ 
            management * # interaction
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
              
            ti(soilt_avg, vwc_avg, by = as.factor(management), k = c(8, 8)) +
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

AIC(r1, r1.2, r1.3)
# r1	  71.64893	4954.354		
# r1.2	91.29666	4921.280		
# r1.3	103.99240	4815.726		

anova(r1.2, r1.3)
# 1    1500.0     1802.0                                     
# 2    1486.3     1660.8 13.653   141.27 9.4429 < 2.2e-16 ***

# same model structure as for daily average NEE
```

Assess model performance:

```{r Reco_GAM_modviz}
appraise(r1.3)
# residuals are distributed normally and the observed values align with the fitted ones
```

```{r Reco_GAM_output}
summary(r1.3)

# R-sq.(adj) =  0.873   Deviance explained =   88%
# -REML = 2475.9  Scale est. = 1.0957    n = 1604
```

##### Interpretation, linear variables:

Note: baseline comparison to conventional management and mature crop stage (estimated Reco = 3.4)

Organic management has a significant effect on Reco baseline (conventional, mature) overall of about 6 units. The crop stage also has significant effects on Reco -- vegetative, reproductive, grain fill, and fallow. The interactions between crop stage and management is also significant, for early, grain fill, and fallow, and trending that way for reproductive Again, the interaction coefficient tells us how much more or less the difference between that crop stage, and the baseline (mature), changes under organic compared to conventional management

##### Interpretaton, smoothing variables:

all smoothing variables are significant in their effects on Reco, and much like NEE, organic management tends toward more nonlinearity in each variable's effects on Reco.

##### Takeaways:

Again and similar to NEE, organic management has a generally significant effect; in this case, in elevating ecosystem respiration across the board. Under conventional (baseline) management, the vegetative, reproductive, and grain fill stages also show significantly higher Reco, indicating that Reco tends to increase as crops develop and peaks at grain fill. In terms of interactions between management and crop stage, there is evidence that the increase in respiration at grain fill is dampened under organic management compared to the same stage under conventional management; so while respiration increases in organic generally, the 'burst' of respiration that occurs around grain fill is higher in the conventional field than in the organic one. Generally speaking, there are nonlinear and greater-in-complexity smoothing variables that represent edaphic characteristics in the organic field, indicating again greater complexity in the soil environment, microbiome, and/or the plant-soil interactions there.

##### Visualize:

###### Interactions, Reco daily average:

```{r Reco_GAM_interactionsplot}

# Predict fitted values with standard errors
pred <- predict(r1.3, newdata, se.fit = TRUE, type = "response")

# using same "new data" as above:
newdata <- newdata %>%
  mutate(
    fit = pred$fit,
    se = pred$se.fit,
    lower = fit - 1.96 * se,
    upper = fit + 1.96 * se
  )

# Plot predicted NEE by crop stage and management
ggplot(newdata, aes(x = crop_stage_simple, y = fit, color = management, group = management)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_line(position = position_dodge(width = 0.5), size = 1) +
  labs(
    title = "Predicted Ecosystem Respiration by Crop Stage and Management Type",
    x = "",
    y = "Predicted Ecos. Resp.",
    color = "Management"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### GPP

#### GAM model development: daily average:

```{r GPPmod_development}

p1 <- gam(GPP ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# plus interaction of vwc and soil temp:
p1.2 <- gam(GPP ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
            
            ti(soilt_avg, vwc_avg, by = as.factor(management), k = c(8, 8)) +
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

# add interaction term between crop stage and management:
p1.3 <- gam(GPP ~ 
            management * # interaction
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(doy, bs = "cc", k = 20) +  # seasonal pattern, cyclical
              
            ti(soilt_avg, vwc_avg, by = as.factor(management), k = c(8, 8)) +
            
            s(soilt_avg, by = as.factor(management), k = 10) +
            s(VPD, by = as.factor(management), k = 10) +
            s(vwc_avg, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = ec_daily_mngmnt4,
          method = "REML")

AIC(p1, p1.2, p1.3)
# p1	  62.60909	7665.234		
# p1.2	80.91281	7627.779		
# p1.3	93.80689	7628.539				

anova(p1.2, p1.3)
#   Resid. Df Resid. Dev     Df Deviance      F   Pr(>F)   
# 1    1511.8     9867.1                                 
# 2    1498.2     9714.3 13.635   152.77 1.7579 0.04144 *

# same model structure as for daily average NEE, daily average Reco
```

Assess model performance:

```{r}
appraise(p1.3)
# residuals are distributed normally and the observed values align with the fitted ones
```

```{r}
summary(p1.3)

# R-sq.(adj) =   0.77   Deviance explained = 78.1%
# -REML = 3848.7  Scale est. = 6.3735    n = 1604
```

#### Interpretation, linear variables:

Baseline GPP (mature crop stage, conventional management) estimate: 3.5

Organic management increases GPP by about 8.9 units in mature crops. Under conventional management, there is significantly higher GPP in the reproductive and grain fill stage (also makes sense). There is significantly lower GPP in the fallow phase (also expected); like respiration, GPP peaks in the grain fill stage. Most interaction terms aren't significant, except that between organic management and vegetative stage, which has a -1.6 unit increase in GPP compared to baseline, indicating that the positive effect of organic on GPP is higher during vegetative crop stages. This might be due in part to the categorization of spring growth (rapid) being 'vegetative' for cover crops and alfalfa that overwinter (aka resuming vegetative growth after dormancy).

#### Interpretation, smoothing variables:

Soil temperature is significant and very complex in the conventional field but not organic, while soil moisture is significant and complex in both fields but more so in the organic field. VPD is significant and nonlinear in the organic and conventional fields as well, and days since tillage is significant in both fields, and more complex in the organic field.

#### Takeaways:

As for NEE and Reco, organic management is highly significant and results in higher baseline GPP (and thus productivity). GPP is strongly moderated by crop stage, particularly grain fill (peak productivity) and dormant (low). An interesting note is that there is no significant interaction between crop stage and management except for dormancy in the organic field, when dormant cover crops over the winter months have greater GPP than those dormant cover crops in the conventional field.

The conventional field has strong nonlinear effects of soil temperature as well as the interaction between soil temperature and moisture, while the organic field does not; however, the organic field demonstrates more complex effects of the other edaphic drivers on GPP including soil moisture. As predicted and expected, seasonality and year-to-year variation are both significant drivers of GPP in all cases.

#### Visualize:

###### Interactions, GPP daily average:

```{r}
# Predict fitted values with standard errors
pred <- predict(p1.3, newdata, se.fit = TRUE, type = "response")

# using same "new data" as above:
newdata <- newdata %>%
  mutate(
    fit = pred$fit,
    se = pred$se.fit,
    lower = fit - 1.96 * se,
    upper = fit + 1.96 * se
  )

# Plot predicted NEE by crop stage and management
ggplot(newdata, aes(x = crop_stage_simple, y = fit, color = management, group = management)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_line(position = position_dodge(width = 0.5), size = 1) +
  labs(
    title = "Predicted Gross Primary Productivity by Crop Stage and Management Type",
    x = "Crop Stage",
    y = "Predicted GPP",
    color = "Management"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## GAM: chamber data, daily averages

```{r}
# again, added in the cover crop data manually along with the alfalfa data (crop stage).
chamber_daily_mngmnt2 <- read_csv("ch_daily_mngmnt_esf.csv")

# make sure management is in factor form
chamber_daily_mngmnt2$management <- factor(as.character(chamber_daily_mngmnt2$management))
# same for crop stage:
chamber_daily_mngmnt2$crop_stage_simple <- factor(
  chamber_daily_mngmnt2$crop_stage_simple,
  levels = c("mature", "early", "vegetative", "reproductive", "grain_fill", "fallow")
)
```

#### NEE: daily average

There are significantly fewer observations with the survey chambers, but in general we want to see agreement with the trends we saw in the eddy covariance dataset (at least for the years for which we have concurrent data: 2019 and 2020). Our daily averages are going to be based on two daytime observations, rather than 24hr averages. Thus the NEE and Rs values will be daytime only and perhaps a bit easier to interpret.

```{r}
#| output = FALSE

# which of these variables co-vary enough to reduce?
cor_vars <- chamber_daily_mngmnt2 %>%
     select(airtempC_mean, soilm_perc_mean, spad_mean, NDVI_mean) %>%
     na.omit()

cor_matrix <- cor(cor_vars)
corrplot(cor_matrix, method = "number", type = "upper")

# test for correlation between SPAD and/or NDVI and crop stage:
# ANOVA
anova_result <- aov(chamber_daily_mngmnt2$spad_mean ~ chamber_daily_mngmnt2$crop_stage_simple)
summary(anova_result)
#                                          Df Sum Sq Mean Sq F value   Pr(>F)    
# chamber_daily_mngmnt2$crop_stage_simple  5   1797   359.4   6.269 0.000144 ***
# Residuals                               49   2809    57.3                                  
anova_result <- aov(chamber_daily_mngmnt2$NDVI_mean ~ chamber_daily_mngmnt2$crop_stage_simple)
summary(anova_result)
#                                         Df Sum Sq Mean Sq F value   Pr(>F)    
# chamber_daily_mngmnt2$crop_stage_simple  5  1.546 0.30918   9.463 7.36e-07 ***
# Residuals                               67  2.189 0.03267                     

# Calculate Eta squared
library(lsr)  # install.packages("lsr") if needed
etaSquared(anova_result)
#                                          eta.sq eta.sq.part
# chamber_daily_mngmnt$crop_stage_simple 0.390129    0.390129
# HUGELY correlated for both SPAD and NDVI. We can use crop stage as a way to measure phenology/greenness bc we have a full column of data with no NAs.

```

Model development:

```{r}
# model 1:
c1 <- gam(NEE_umolm2sec_mean ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(month, bs = "cc", k = 6, by = as.factor(management)) +  # seasonal pattern

            s(airtempC_mean, by = as.factor(management), k = 10) +
            s(soilm_perc_mean, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = chamber_daily_mngmnt2,
          method = "REML")

# add interaction term between crop stage and management:
c2 <- gam(NEE_umolm2sec_mean ~ 
            management *
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(month, bs = "cc", k = 6, by = as.factor(management)) +  # seasonal pattern
            
            s(airtempC_mean, by = as.factor(management), k = 10) +
            s(soilm_perc_mean, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = chamber_daily_mngmnt2,
          method = "REML")

# add tensor interaction between air temp and soil moisture
c3 <- gam(NEE_umolm2sec_mean ~ 
            management *
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(month, bs = "cc", k = 6, by = as.factor(management)) +  # seasonal pattern
            
            ti(soilm_perc_mean, airtempC_mean, by = as.factor(management), k = c(8, 8)) +

            s(airtempC_mean, by = as.factor(management), k = 10) +
            s(soilm_perc_mean, by = as.factor(management), k = 10) +
            s(days_since_tillage, by = as.factor(management), k = 8),

          data = chamber_daily_mngmnt2,
          method = "REML")
```

```{r}
# compare using AIC values:
AIC(c1, c2, c3)
#         df      AIC
# c1	18.16276	571.5739		
# c2	24.89599	567.2237		
# c3	32.87916	558.5610	

anova(c1, c2)
#   Resid. Df Resid. Dev    Df Deviance      F  Pr(>F)  
# 1    79.459     1351.3                                
# 2    72.437     1126.7 7.022   224.63 2.1514 0.04853 *

anova(c2, c3)
# 1    72.437    1126.65                                 
# 2    62.424     876.28 10.013   250.37 1.9923 0.04917 *

# Check diagnostics of final model:
appraise(c3) #looks pretty good
```

```{r}
summary(c3)
```

##### Interpretation:

While there is no significant effect of management on chamber-based daily average NEE, there it is clear that both crop stage (reproductive, grain fill) both reduce NEE when compared to the baseline of conventional, mature crop by quite a bit. Additionally, the interaction between management and crop stage is significant, with enhanced effects of early, vegetative, reproductive, and grain fill stages on NEE in the organic field compared to the conventional one.

The only smoothing factors that are significant in driving NEE are days since tillage, though the effect is much more complex in the organic field than the conventional one.

##### Takeaways:

These results largely align with those data collected from the EC towers, reinforcing the conclusions that organic management generally reduces NEE but that this effect varies by crop stage. The drivers of the effect are much less linear / more complex in the organic field compared to the conventional one.

##### Visualize:

###### Predicted Through Crop Stage:

```{r}
# Create prediction data
pred_data <- expand.grid(
  crop_stage_simple = factor(c("mature", "early", "vegetative", "reproductive", 
                                "grain_fill", "fallow")),
  management = factor(c("conventional", "organic")),
  # Hold other variables at representative values
  year = median(chamber_daily_mngmnt2$year),
  month = median(chamber_daily_mngmnt2$month),
  airtempC_mean = mean(median(chamber_daily_mngmnt2$airtempC_mean), na.rm = TRUE),
  soilm_perc_mean = mean(chamber_daily_mngmnt2$soilm_perc_mean, na.rm = TRUE),
  days_since_tillage = mean(chamber_daily_mngmnt2$days_since_tillage, na.rm = TRUE)
)

# Get predictions with confidence intervals
pred_data$fit <- predict(c3, newdata = pred_data, se.fit = TRUE)$fit
pred_data$se <- predict(c3, newdata = pred_data, se.fit = TRUE)$se.fit

# Plot
ggplot(pred_data, aes(x = crop_stage_simple, y = fit, 
                      color = management, group = management)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se, 
                  fill = management), alpha = 0.2) +
  geom_point(size = 3) +
  labs(title = "NEE Through Crop Development",
       x = "Crop Stage", y = "predicted NEE") +
  theme_bw()
```

###### Predicted vs. Observed

```{r}
# Add predictions to original data
chamber_daily_mngmnt2$predictedNEE <- predict(c3)

chamber_daily_mngmnt2 %>% 
  ggplot(aes(x = predictedNEE, y = NEE_umolm2sec_mean)) +
  geom_point(aes(color = management, shape = crop_stage_simple), 
             alpha = 0.75, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~management) +
  labs(title = "GAM Model Fit: Observed vs. Predicted NEE",
       x = "Predicted NEE",
       y = "Observed NEE") +
  theme_bw()
```

### Reco: daily average

```{r}
# simple:
n1 <- gam(Rs_umolm2sec_mean ~ 
            management +
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(month, bs = "cc", k = 6, by = as.factor(management)) +  # seasonal pattern
            
            s(soilm_perc_mean, k = 10, by = as.factor(management)) +
            s(airtempC_mean, k = 10, by = as.factor(management)) +
            s(days_since_tillage, k = 8, by = as.factor(management)),

          data = chamber_daily_mngmnt2,
          method = "REML")

# with interaction of management and crop stage:
n2 <- gam(Rs_umolm2sec_mean ~ 
            management *
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(month, bs = "cc", k = 6, by = as.factor(management)) +  # seasonal pattern
            
            s(soilm_perc_mean, k = 10, by = as.factor(management)) +
            s(airtempC_mean, k = 10, by = as.factor(management)) +
            s(days_since_tillage, k = 8, by = as.factor(management)),

          data = chamber_daily_mngmnt2,
          method = "REML")

# add tensor interaction between soil moisture and air temp:
n3 <- gam(Rs_umolm2sec_mean ~ 
            management *
            crop_stage_simple +
            s(year, bs = "re", by = as.factor(management)) + # random effect of year
            s(month, bs = "cc", k = 6, by = as.factor(management)) +  # seasonal pattern
            
            ti(soilm_perc_mean, airtempC_mean, by = as.factor(management), k = c(8, 8)) +
         
            s(soilm_perc_mean, k = 10, by = as.factor(management)) +
            s(airtempC_mean, k = 10, by = as.factor(management)) +
            s(days_since_tillage, k = 8, by = as.factor(management)),

          data = chamber_daily_mngmnt2,
          method = "REML")

AIC(n1, n2, n3)
#       df        AIC
# n1	25.25305	538.8713		
# n2	29.72541	507.1832		
# n3	33.21324	508.7876	

anova(n1, n2)
#   Resid. Df Resid. Dev     Df Deviance      F    Pr(>F)    
# 1    72.239     837.49                                     
# 2    67.726     553.24 4.5126   284.25 8.0638 1.034e-05 ***

anova(n2, n3)
#   Resid. Df Resid. Dev     Df Deviance     F Pr(>F)
# 1    67.726     553.24                             
# 2    63.401     523.73 4.3252   29.506 0.888 0.4828

# going to keep the tensor interaction because the logic holds for this analysis as well as for the others: there is a likelihood of the air temp (in lieu of soil temp, here) co-varying with soil moisture in a way that is interactive, even if the model comparison shows a slightly higher AIC.  The difference is extremely slight, and model n3 explains a significant amount of the variance in the dataset (84.8% in n3 versus 85% in n2).
```

```{r}
appraise(n3) #looks good
```

```{r}
summary(n3)
```

##### Interpretation:

Again, management does not have a clear significant effect when holding all else equal. However, crop stage once again indicates differences in ecosystem respiration with fallow having significantly higher respiration when compared to conventional mature, and grain fill indicating higher respiration as well. The interaction between management and crop stage is significant for the reproductive and grain fill stages, indicating a much lower respiration rate in these stages when compared to conventional mature.

As for chamber-based NEE, chamber-based ecosystem respiraiton is significantly moderated by days since tillage and once again, the effect is far more complex for the organic field than the conventional field. In addition, the organic field has a significant effect of air temperature, which is fairly linear in its positive effect, and seasonality (here, month) indicates a slightly positive effect in the organic field as well.

##### Takeaways:

Again this analysis largely supports the conclusions from the EC tower data, and days since tillage seems to be the most prominent smoothing variable in both fields that emerges from the "all smooths are significant" outcomes of the tower data, in terms of moderating both NEE and ecosystem respiration on the scale of the collars.

##### Visualize:

###### Predicted Through Crop Stage:

```{r}
# Use existing prediction data

# Get predictions with confidence intervals
pred_data$fit <- predict(n3, newdata = pred_data, se.fit = TRUE)$fit
pred_data$se <- predict(n3, newdata = pred_data, se.fit = TRUE)$se.fit

# Plot
ggplot(pred_data, aes(x = crop_stage_simple, y = fit, 
                      color = management, group = management)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se, 
                  fill = management), alpha = 0.2) +
  geom_point(size = 3) +
  labs(title = "Respiration Through Crop Development",
       x = "Crop Stage", y = "Predicted R_s") +
  theme_bw()
```

###### Predicted vs. Observed

```{r}
# Add predictions to original data
chamber_daily_mngmnt2$predictedRs <- predict(n3)

chamber_daily_mngmnt2 %>% 
  ggplot(aes(x = predictedRs, y = Rs_umolm2sec_mean)) +
  geom_point(aes(color = management, shape = crop_stage_simple), 
             alpha = 0.75, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~management) +
  labs(title = "GAM Model Fit: Observed vs. Predicted Respiration",
       x = "Predicted Rs",
       y = "Observed Rs") +
  theme_bw()
```
